---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


# Loading and preprocessing the data

### Load libraries
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
```

### Unzip and Load the data file  
### Read the file activity.csv into an R data frame  
### Preprocess the data to summarize the number of steps each day.
```{r preprocess}
zipfile <- "activity.zip"
exdir = "data"
unzip(zipfile, exdir=exdir)
infile <- "data/activity.csv"
df <- read.csv(infile)
df2 <- df %>% mutate (converted_date = ymd(date), na_steps = is.na(steps), ID = row_number())
df3 <- df2 %>% group_by( converted_date ) %>% summarize(daily_steps = sum(steps), missing_steps = sum(na_steps))
```

# What is mean total number of steps taken per day?


### Show the mean number of daily steps
```{r}
original_mean <- mean(df3$daily_steps, na.rm = TRUE )
original_mean
```

### SHow the median for daily steps
```{r}
original_median <- median(df3$daily_steps, na.rm = TRUE )
original_median
```

### Show the number of steps each day in a bar chart
```{r bar_chart_steps_per_day}
ggplot(df3, mapping=aes(x=converted_date, y=daily_steps)) + 
  geom_bar(stat = "identity") +
  labs(title="Number of Steps Each Day")
```


### show a histogram of frequency values for daily steps
```{r histogram_of_daily_steps}
hist(df3$daily_steps,breaks=10)
```


# What is the average daily activity pattern?

### Group steps by interval, so we can compute the mean and sum for each interval
```{r}
steps_by_interval <- df2 %>% group_by ( interval ) %>% 
  summarize (sum_steps = sum(steps, na.rm=TRUE), 
             mean_steps = mean(steps, na.rm=TRUE),
             average_steps_integer = as.integer(round(mean_steps, 0)),
             count_steps = sum(!is.na(steps)))
```

### Plot the timeline of average steps versus interval, averaged across all days
```{r timeline_of_average_steps_per_interval}
g <- ggplot(steps_by_interval, mapping=aes(x=interval, y=mean_steps)) + geom_line()
g <- g + xlab("interval -- time of day (HH:MM)")
g <- g + ylab("Average steps per interval across all days")
g <- g + ggtitle("Timeline of Average Steps per Interval")
g
```

### Determine the maximum average step value across intervals.
```{r}
max_steps <- max( steps_by_interval$mean_steps)
max_steps
```

```{r}
### Determine which time interval has the max_steps value.
max_steps_interval <- steps_by_interval[steps_by_interval$mean_steps == max_steps, ]$interval
max_steps_interval
```

## Apparently the largest number of average steps 206.1698 occurred at interval
## 835 or 8:35 a.m.


# Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset
### (i.e. the total number of rows with `NA`s)
```{r}
summary(df2)
```

### The summary function shows that 2304 values of steps are NA.
```{r}
dates_with_na_steps <- df2 %>% filter(is.na(steps)) %>% 
  group_by(date) %>% summarize(count_of_nas =n())
dates_with_na_steps
```

### It turns out that there are 288 5-minute intervals each day,
### and we are missing all of the steps data on 8 full days.
### This accounts for all of the 2304 = 8 * 288 missing values.

## 2. Devise a strategy for filling in all of the missing values in the dataset. 
## The strategy does not need to be sophisticated. For example, 
## you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
## 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

### Create a function to replace a missing step value
### with the corresponding average for that 5 minute interval
### across all the days with data.
```{r}
impute_step_value <- function( steps, interval ) {
  if(!is.na(steps)) {
    ## just return the same steps value where data is present.
    result <- steps
  } else {
    ## interval is an integer in 24 hour time format HH:MM
    ## We want to find an index based on the number of 5 minute intervals
    ## within the current day.
    index <- 1 + 12 * (interval %/% 100) + (interval %% 60) %/% 5
    ## for missing values return the average for that time of day
    ## (rounded to an intger).
    result <- steps_by_interval$average_steps_integer[index]
  }
  if(is.na(result)) {
    print(paste('interval: ', interval))
  }
  result
}
```

### create a function to call impute_step_value for
### each row of data.frame df2
```{r}
impute_row <- function(row_index) {
  impute_step_value(df2$steps[row_index], df2$interval[row_index])
}
```

### Create a vector of the original step values merged with imputed step values.
```{r}
steps_imputed <- sapply(1:nrow(df2), impute_row)
```

### Form a new data.frame with selected columns  and including the imputed step values.
```{r}
imputed_steps_df <- df2 %>% select( date, interval, converted_date ) %>% cbind(steps_imputed)
```

### Group the data by date to sum total steps per day
```{r}
daily_df <- imputed_steps_df %>% group_by( converted_date ) %>% summarize(daily_steps = sum(steps_imputed))
```

## 4. Make a histogram of the total number of steps taken each day and Calculate and 
## report the **mean** and **median** total number of steps taken per day. 
## Do these values differ from the estimates from the first part of the assignment? 
## What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r histogram_with_imputed_steps} 
hist(daily_df$daily_steps,breaks=10)
```

### Compare mean and median after imputing steps to the orginal values.
```{r}
mean(daily_df$daily_steps)
original_mean
median(daily_df$daily_steps)
original_median
```
### The means and medians are close but do not exactly match. This is reasonable
### because we have added some estimated data.

# Are there differences in activity patterns between weekdays and weekends?

## For this part the `weekdays()` function may be of some help here. Use
## the dataset with the filled-in missing values for this part.

## 1. Create a new factor variable in the dataset with two levels -- 
## "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

## Define a new function to detect weekend days from day names.
## Then categorize each row of the data set as weekday or weekend.
```{r}
weekends <- function(day_of_week) {
  if("Saturday" == day_of_week || "Sunday" == day_of_week)  {
    weekend <- "weekend"
  } else {
    weekend <- "weekday"
  }
  weekend
}

new_dataset <- imputed_steps_df %>% 
  mutate(day_of_week=weekdays(converted_date))

day_type <- sapply(new_dataset$day_of_week, weekends)

new_dataset <- cbind(new_dataset, day_type)
  
str(new_dataset)
summary(new_dataset)
```

## Make a panel plot containing a time series plot (i.e. `type = "l"`) of the
## 5-minute interval (x-axis) and the average number of steps taken, 
## averaged across all weekday days or weekend days (y-axis).

### Prepare the data, grouping by interval and weekday vs. weekend
```{r}
steps_by_interval2 <- new_dataset %>% group_by ( interval, day_type ) %>% 
  summarize (sum_steps = sum(steps_imputed), 
             mean_steps = mean(steps_imputed))
```



### Plot the timeline of average steps versus interval, averaged across all days
### with weekdays versus weekends in separate panels
```{r weekdays_vs_weekends}
g <- ggplot(steps_by_interval2, mapping=aes(x=interval, y=mean_steps)) + geom_line()
g <- g + facet_grid(day_type ~ .)
g <- g + xlab("interval -- time of day (HH:MM)")
g <- g + ylab("Average steps per interval across all days")
g <- g + ggtitle("Timeline of Average Steps per Interval (Weekdays vs. Weekends)")
g
```

